### 2022-05-21 main/chapter_2ljets.tex
add metsig and mt2 descriptions

We use two signal models in the $\twoljets$ electroweak analysis.
One is an MSSM chargino-neutralino model (C1N2) which acts to produce
$\chargino{1}\textrm{--}\neutralino{2}$ pairs, which decay via
$W\rightarrow q\bar q$ and
$Z\rightarrow \ell^\pm \ell^\mp$ respectively..
The other is a GMSB model which, after a variety of initial states produces
$\neutralino{1}\textrm{--}\neutralino{1}$ pairs.
These neutralinos decay,
via Higgs and $Z$ bosons in variable branching fractions,
to final states again containing $\ell^\pm \ell^\mp$ and jet-seeding partons.
Diagrams illustrating these C1N2 and GMSB signal models are shown in
Figure~\ref{fig:2ljets_signal_diagrams}.

@@

\begin{figure}[t]
\centering
\includegraphics[width=0.48\textwidth]{figures/2ljets_c1n2_llqqn1n1_wz.pdf}
\hfill
\includegraphics[width=0.45\textwidth]{figures/2ljets_n1n1_hhggzz.pdf}
\caption{%
Supersymmetric signal processes in the $\twoljets$ electroweak
analysis~\cite{atlas2022searches}.
\\[0.5em]
Left: C1N2, where the initial $\chargino{1}\textrm{--}\neutralino{2}$ pair
is produced through a $s$-channel $W^{\pm}$ resonance and the masses of
weak bosons in decays are bounded by the mass splitting
$m(\chargino{1}, \neutralino{2}) - m(\neutralino{1})$.
We explore the parameters
$m(\chargino{1}, \neutralino{2})$ and $m(\neutralino{1})$.
\\[0.5em]
Right: GMSB, where the initial $\neutralino{1}\textrm{--}\neutralino{1}$ pair
is produced by soft decays from pairs including $\chargino{1}$,
$\neutralino{2}$ or $\neutralino{1}$.
Although the $h$ and $Z$ bosons decay to many states, we target
$Z\rightarrow \ell\ell$ with
$h/Z\rightarrow bb/jj$.
We explore the parameters
$m(\neutralino{1})$ and $B(\neutralino{1} \rightarrow h \tilde{G})$ with fixed
$m(\chargino{1}, \neutralino{2}) - m(\neutralino{1}) = 1~\eV[G]$.
}
\label{fig:2ljets_signal_diagrams}
\end{figure}

@@

\subsection{$\met$ significance}
\label{sec:metsig}
It scales $\met$ by an uncertainty derived from the resolutions of
reconstructed objects~\cite{atlas_met_significance}.

\subsection{Stranverse mass $\mttwo$}
\label{sec:mt2}
% angle between the vector sum of the lepton pair and the missing transverse momentum.
% leptons and the missing transverse momentum.
% If two massive particles are pair produced and each decays to a
% lepton and something massless and invisible,
% then this estimates a greatest lower bound on the invariant
% mass of the pair produced particles.
% $\mttwo(\ell, \ell; 0)$ is useful for excluding standard model backgrounds;
% di-leptonic ttbar ($tt\rightarrow WW\rightarrow\ell\nu\ell\nu$)
% and $WW\rightarrow\ell\nu\ell\nu$ have a kinematic endpoint
% at the $W$ pole mass, and $\tau\tau$ is constrained within measurement error of zero.

@@

\clearpage
\subsection{Cheat-sheet}
Uncomfortably many event variables are used to define the regions of this
analysis.
To aid future revision of their meanings, definitions and references for these
variables are collected here.
\begin{itemize}
\item $\met$: missing transverse momentum
\item $\metsig$: object-based missing transverse momentum significance;\\
described in Section~\ref{sec:metsig}
\item $\ptjone$: transverse momentum of the hardest jet%
\vspace{0.5em}
\item $\mll$: invariant mass of the hardest two leptons
\item $\mjj$: invariant mass of the two hardest jets
\item $\mbb$: invariant mass of the two hardest b-tagged jets
\item $\mjetone$: mass of the hardest jet
\item $\mttwoll$: `stransverse` mass of the lepton pair with massless invisibles;\\
described in Section~\ref{sec:mt2}
\vspace{0.5em}
\item $\rll$: distance in $\eta\textrm{--}\phi$ between the lepton pair
\item $\rjj$: distance in $\eta\textrm{--}\phi$ between the two hardest jets%
\vspace{0.5em}
\item $\njet$: number of signal jets
\item $\nbtag$: number of b-tagged jets;
described in Section~\ref{sec:btagging}
\vspace{0.5em}
\item $\dphillmet$: azimuthal angle between the lepton pair and $\ptmiss$
\item $\dphijmet$: azimuthal angle between the hardest jet and $\ptmiss$
\end{itemize}
All of these are physical properties of particles or collections of particles.
Please remember that none is an exact or true value;
all are are estimates based on approximate reconstructions.
While this distinction is easy to drop in lax language, is necessary to
understand the shapes of noisy variables such as $\mjj$ and indeed the
existence of $\metsig$.

>>>>

We use two signal models in the $\twoljets$ electroweak analysis.
One is an MSSM chargino-neutralino model (C1N2) which acts to produce
$\chargino_1\textrm{--}\neutralino_2$ pairs, which decay via
$W\rightarrow q\bar q$ and
$Z\rightarrow \ell^\pm \ell^\mp$ respectively..
The other is a GMSB model which, after a variety of initial states produces
$\neutralino_1\textrm{--}\neutralino_1$ pairs.
These neutralinos decay,
via Higgs and $Z$ bosons in variable branching fractions,
to final states again containing $\ell^\pm \ell^\mp$ and jet-seeding partons.
Diagrams illustrating these C1N2 and GMSB signal models are shown in
Figure~\ref{fig:2ljets_signal_diagrams}.

@@

\begin{figure}[t]
\centering
\includegraphics[width=0.48\textwidth]{figures/2ljets_c1n2_llqqn1n1_wz.pdf}
\hfill
\includegraphics[width=0.45\textwidth]{figures/2ljets_n1n1_hhggzz.pdf}
\caption{%
Supersymmetric signal processes in the $\twoljets$ electroweak
analysis~\cite{atlas2022searches}.
\\[0.5em]
Left: C1N2, where the initial $\chargino_1\textrm{--}\neutralino_2$ pair
is produced through a $s$-channel $W^{\pm}$ resonance and the masses of
weak bosons in decays are bounded by the mass splitting
$m(\chargino_1, \neutralino_2) - m(\neutralino_1)$.
We explore the parameters
$m(\chargino_1, \neutralino_2)$ and $m(\neutralino_1)$.
\\[0.5em]
Right: GMSB, where the initial $\neutralino_1\textrm{--}\neutralino_1$ pair
is produced by soft decays from pairs including $\chargino_1$,
$\neutralino_2$ or $\neutralino_1$.
Although the $h$ and $Z$ bosons decay to many states, we target
$Z\rightarrow \ell\ell$ with
$h/Z\rightarrow bb/jj$.
We explore the parameters
$m(\neutralino_1)$ and $B(\neutralino_1 \rightarrow h \tilde{G})$ with fixed
$m(\chargino_1, \neutralino_2) - m(\neutralino_1) = 1~\eV[G]$.
}
\label{fig:2ljets_signal_diagrams}
\end{figure}

@@

\subsection{$\met$ significance}
\label{sec:metsig}

Most common Standard Model processes at the LHC do not produce hard
invisible particles.
Our supersymmetric signals, however, do.
This makes the missing transverse momentum $\ptmiss$ its magnitude $\met$
important features for separating signal and background events in all parts
of the $\twoljets$ search.

The experimental resolution of $\ptmiss$, however, is imprecise.
Importantly also, the uncertainty on any inferred $\ptmiss$ varies
significantly from one event to another.

A muon, for example, will tend to have its momentum measured more precisely
than a jet of similar energy.
Unless, as discussed in Chapter~\ref{chapter:experiment}, the muon is hard
enough that its curvature through \atlas' magnetic fields is poorly tracked.
Unless, in the right circumstances, that jet originates from a $b$ quark,
such that the weak decays of its hadrons tend to lose momentum in
neutrinos or soft muons. (The $\metsig$ variable used in this analysis is not
aware of $b$-tagging information, but perhaps it could benefit from some.)

By incorporating an uncertainty in $\ptmiss$ tailored to each event, the
variable $\metsig$ helps to separate those events with real and well-measured
$\met$ from those where the measures $\ptmiss$ is more likely to result from
mismeasurements.

In the context of $\twoljets$ selections, $\metsig$ suppresses $t\bar t$
contributions more than $\met$ alone, plausibly since $t\bar t$ events are
accompanied with more hadronic activity than other sources of $\ptmiss$.
What remains of the standard model in the high-$\metsig$ tail for $\twoljets$
selections is almost pure diboson -- $ZZ$ with a hard $Z\rightarrow \nu\nu$,
and $W^\pm Z$ where the charged lepton in $W^\pm\rightarrow\ell^\pm\nu$
is not reconstructed.
\TODO{plot of diboson decomposition here?}
These effects are demonstrated from simulations in
Figure~\ref{fig:2ljets_presel_met}.

Uncertainty in $\ptmiss$ is modelled for $\metsig$ by a covariance matrix, $S$,
in the transverse plane, and used to define
\begin{equation}
\metsig
=
\ptmiss S^{-1} \ptmiss
,
\end{equation}
which can be rearranged into a ratio of $(\met)^2$ to a function of elements
of the covariance matrix~\cite{atlas_met_significance}.
This is the square of a Mahalanobis (standard-deviations) distance between
the origin and $\ptmiss$~\cite{mahalanobis1936generalised}.

Just as $\ptmiss$ is constructed from a (negative) sum of contributions from
objects in an event,
$S$ is a sum of covariances assigned to those objects;
both means and variances add linearly.

\subsection{Stranverse mass $\mttwo$}
\label{sec:mt2}
% TODO
% angle between the vector sum of the lepton pair and the missing transverse momentum.
% leptons and the missing transverse momentum.
% If two massive particles are pair produced and each decays to a
% lepton and something massless and invisible,
% then this estimates a greatest lower bound on the invariant
% mass of the pair produced particles.
% $\mttwo(\ell, \ell; 0)$ is useful for excluding standard model backgrounds;
% di-leptonic ttbar ($tt\rightarrow WW\rightarrow\ell\nu\ell\nu$)
% and $WW\rightarrow\ell\nu\ell\nu$ have a kinematic endpoint
% at the $W$ pole mass, and $\tau\tau$ is constrained within measurement error of zero.

Just as the slepton is a supersymmetric partner to a lepton, the stransverse
mass  $\mttwo$ is a partner to the `transverse mass' $m_T$
(to be defined shortly),
that is useful in searches for supersymmetric particles.

The stransverse mass of an event is a greatest lower bound on the mass of a
pair-produced resonance, that applies when each partner decays semi-invisibly.
For example, an on-shell $\tilde\ell^+\tilde\ell^-$ pair decaying as
$\tilde\ell^+ \rightarrow \ell^+ \neutralino_1
,~
\tilde\ell^- \rightarrow \ell^- \neutralino_1$ has
$\mttwo(\ell^+, \ell^-, m(\neutralino_1)) \leq m(\tilde\ell)$, plus some error
margin for experimental resolution.
Such cases are common since, R-parity conservation implies pair production,
and the idea that dark matter could comprise the lightest supersymmetric
particle motivates invisible decays.

Lacking direct observation of sleptons, $\mttwo$ remains useful in separating
standard model processes; dileptonic $t\bar t$ and $W^\pm W^\mp$ decays both
satisfy $\mttwo(\ell^+, \ell^-, 0) \leq m(W)$ (since $0 \approx m(\nu)$).
The $\twoljets$ electroweak search depends on this effect;
most selections require $\mttwo(\ell^+, \ell^-, 0) > 80~\eV[G]$, or similar,
to reduce $t\bar t$, $W^\pm W^\mp$ and $\tau^\pm\tau^\mp$ backgrounds.

To be precise, the stransverse mass explores how to split the observed
$\ptmiss$ between the two visible decay products in an event, and can be
defined as
\begin{equation}
\mttwo(a_\mu, b_\mu, m)
=
\min_{\vec p\,+\,\vec q\,=\,\ptmiss}
\max
\begin{Bmatrix}
m_T(a_\mu + \{\!\sqrt{m^2 - \vec p_T^{\:2}},\,\vec p\,\}_\mu)\\
m_T(b_\mu + \{\!\sqrt{m^2 - \vec q_T^{\:2}},\,\vec q\,\}_\mu)\\
\end{Bmatrix}
,
\end{equation}
in which $m_T^2(p_\mu) = p_0^2 - p_1^2 - p_2^2$~\cite{lester1999measuring}.
As an extension, one can assign different masses $m_p$ and $m_q$ to the
the two invisible products~\cite{lester2015bisection}.

Numerical evaluation of $\mttwo$ can be performed efficiently with interval
bisection algorithms, which seek the least resonance mass for which the
transverse masses allow any compatible missing momentum
assignment~\cite{cheng2008minimal, lester2015bisection}.
Public code for calculating $\mttwo$ is available as a Python
library~\cite{gillam2021mt2}, to which I contributed a faster and more robust
implementation of the bisection algorithm by
Lester and Nachman~\cite{lester2015bisection} during this project.

@@

\subsection{Cheat-sheet}
Uncomfortably many event variables are used to define the regions of this
analysis.
To aid future revision of their meanings, definitions and references for these
variables are collected here.
\begin{itemize}
\item $\met$: missing transverse momentum
\item $\metsig$: object-based missing transverse momentum significance;\\
described in Section~\ref{sec:metsig}
\item $\ptjone$: transverse momentum of the hardest jet%
\vspace{0.5em}
\item $\mll$: invariant mass of the hardest two leptons
\item $\mjj$: invariant mass of the two hardest jets
\item $\mbb$: invariant mass of the two hardest b-tagged jets
\item $\mjetone$: mass of the hardest jet
\item $\mttwoll$: `stransverse` mass of the lepton pair with massless invisibles;\\
described in Section~\ref{sec:mt2}
\vspace{0.5em}
\item $\rll$: distance in $\eta\textrm{--}\phi$ between the lepton pair
\item $\rjj$: distance in $\eta\textrm{--}\phi$ between the two hardest jets%
\vspace{0.5em}
\item $\njet$: number of signal jets
\item $\nbtag$: number of b-tagged jets;
described in Section~\ref{sec:btagging}
\vspace{0.5em}
\item $\dphillmet$: azimuthal angle between the lepton pair and $\ptmiss$
\item $\dphijmet$: azimuthal angle between the hardest jet and $\ptmiss$
\end{itemize}
All of these are physical properties of particles or collections of particles.
Please remember that none is an exact or true value;
all are are estimates based on approximate reconstructions.
While this distinction is easy to drop in lax language, is necessary to
understand the shapes of noisy variables such as $\mjj$ and indeed the
existence of $\metsig$.
